Imputation of Incomplete Multilevel Data with R

### 2023-12-07 docs/manuscript.qmd
Add scatterplots

---
title: "Imputation of Incomplete Multilevel Data with R"
title-formatted: "Imputation of Incomplete Multilevel Data with [R]{.proglang}"
format:
    jss-pdf:
        keep-tex: true
        fig-pos: 't'
    jss-html: default
author:
  - name: Hanne I. Oberman
    affiliations:
      - name: Utrecht University
        department: Methodology and Statistics
        address: Padualaan 14
        city: Utrecht
        country: The Netherlands
        postal-code: 3584 CH
      - Journal of Statistical Software
    orcid: 0000-0003-3276-2141
    email: h.i.oberman@uu.nl
    url: https://www.hanneoberman.github.io
  - name: Johanna Muñoz
    affiliations:
      - name: University Medical Center Utrecht
        department: Julius Centre for Health Sciences and Primary Care
        address: Universiteitsweg 100
        city: Utrecht
        country: The Netherlands
        postal-code: 3584 CG
      - Journal of Statistical Software
    orcid: 0000-0002-2384-5415
  - name: Valentijn M.T. de Jong
    affiliations:
      - name: University Medical Center Utrecht
        department: Julius Centre for Health Sciences and Primary Care
        city: Utrecht
        country: The Netherlands
      - name: European Medicines Agency
        department: Data Analytics and Methods Task Force
        city: Amsterdam
        country: The Netherlands
      - Journal of Statistical Software
    orcid: 0000-0001-9921-3468
  - name: Gerko Vink
    affiliations:
      - name: University Medical Center Utrecht
        department: Julius Centre for Health Sciences and Primary Care
        address: Universiteitsweg 100
        city: Utrecht
        country: The Netherlands
        postal-code: 3584 CG
      - Journal of Statistical Software
    orcid: 0000-0001-9767-1924
  - name: Thomas P.A. Debray
    affiliations:
      - name: University Medical Center Utrecht
        department: Julius Centre for Health Sciences and Primary Care
        address: Universiteitsweg 100
        city: Utrecht
        country: The Netherlands
        postal-code: 3584 CG
      - Journal of Statistical Software
    orcid: 0000-0002-1790-2719    
abstract: |
  This tutorial illustrates the imputation of incomplete multilevel data with the [R]{.proglang} packackage [mice]{.pkg}. Our scope is only simple multilevel models, to show how imputation can yield less biased estimates from incomplete clustered data. More complex models can be accomodated, but are outside the scope of this paper. Incomplete multilevel data requires careful consideration of the missing data problem and analysis strategy. In this tutorial, we focus on a popular strategy for accommodating missingness in multilevel data: replacing the missing data with one or more plausible values, i.e., imputation.Imputation separates the missing data problem from the main analysis and the completed data can be analyzed as if it has been fully observed. This tutorial illustrates the imputation of incomplete multilevel data with the statistical programming language R. We aim to show how imputation can yield less biased estimates from incomplete clustered data. We provide practical guidelines and code snippets for different missing data situations, including non-ignorable missingness mechanisms. For brevity, we focus on multilevel imputation using chained equations with the R mice package and its adjacent packages.

@@

```{R}
#| prompt: true
#| warning: false
#| message: false
library(mice)
library(ggmice)
```

@@

```{R}
#| prompt: true
dat <- popmis[, c("school", "popular", "sex")] 
```

@@

```{R}
#| layout-ncol: 2
#| echo: false
plot_pattern(dat)
plot_corr(dat)
```

@@

```{R}
#| prompt: true
meth <- make.method(dat)
```

```{R}
#| prompt: true
pred <- quickpred(dat)
plot_pred(pred, method = meth)
```

```{R}
#| prompt: true
imp <- mice(
  data = dat,
  method = meth,
  predictorMatrix = pred,
  printFlag = FALSE
)
```

```{R}
#| prompt: true
plot_trace(imp)
```

>>>>

---
title: "Imputation of Incomplete Multilevel Data with R"
title-formatted: "Imputation of Incomplete Multilevel Data with [R]{.proglang}"
format:
    jss-pdf:
        keep-tex: true
        fig-pos: 't'
        fig-width: 7
        fig-height: 3.5
    jss-html: default
author:
  - name: Hanne I. Oberman
    affiliations:
      - name: Utrecht University
        department: Methodology and Statistics
        address: Padualaan 14
        city: Utrecht
        country: The Netherlands
        postal-code: 3584 CH
      - Journal of Statistical Software
    orcid: 0000-0003-3276-2141
    email: h.i.oberman@uu.nl
    url: https://www.hanneoberman.github.io
  - name: Johanna Muñoz
    affiliations:
      - name: University Medical Center Utrecht
        department: Julius Centre for Health Sciences and Primary Care
        address: Universiteitsweg 100
        city: Utrecht
        country: The Netherlands
        postal-code: 3584 CG
      - Journal of Statistical Software
    orcid: 0000-0002-2384-5415
  - name: Valentijn M.T. de Jong
    affiliations:
      - name: University Medical Center Utrecht
        department: Julius Centre for Health Sciences and Primary Care
        city: Utrecht
        country: The Netherlands
      - name: European Medicines Agency
        department: Data Analytics and Methods Task Force
        city: Amsterdam
        country: The Netherlands
      - Journal of Statistical Software
    orcid: 0000-0001-9921-3468
  - name: Gerko Vink
    affiliations:
      - name: University Medical Center Utrecht
        department: Julius Centre for Health Sciences and Primary Care
        address: Universiteitsweg 100
        city: Utrecht
        country: The Netherlands
        postal-code: 3584 CG
      - Journal of Statistical Software
    orcid: 0000-0001-9767-1924
  - name: Thomas P.A. Debray
    affiliations:
      - name: University Medical Center Utrecht
        department: Julius Centre for Health Sciences and Primary Care
        address: Universiteitsweg 100
        city: Utrecht
        country: The Netherlands
        postal-code: 3584 CG
      - Journal of Statistical Software
    orcid: 0000-0002-1790-2719    
abstract: |
  This tutorial illustrates the imputation of incomplete multilevel data with the [R]{.proglang} packackage [mice]{.pkg}. Our scope is only simple multilevel models, to show how imputation can yield less biased estimates from incomplete clustered data. More complex models can be accomodated, but are outside the scope of this paper. Incomplete multilevel data requires careful consideration of the missing data problem and analysis strategy. In this tutorial, we focus on a popular strategy for accommodating missingness in multilevel data: replacing the missing data with one or more plausible values, i.e., imputation.Imputation separates the missing data problem from the main analysis and the completed data can be analyzed as if it has been fully observed. This tutorial illustrates the imputation of incomplete multilevel data with the statistical programming language R. We aim to show how imputation can yield less biased estimates from incomplete clustered data. We provide practical guidelines and code snippets for different missing data situations, including non-ignorable missingness mechanisms. For brevity, we focus on multilevel imputation using chained equations with the R mice package and its adjacent packages.

@@

```{R}
#| prompt: true
#| warning: false
#| message: false
set.seed(123)
library(mice)
library(ggmice)
library(ggplot2)
library(miceadds)
library(lme4)
library(mitml)
library(broom.mixed)
```

@@

```{R}
#| prompt: true
dat <- popmis[, c("school", "teachpop", "popular", "texp", "sex")] 
```

```{R}
#| fig-cap: "Polar axis plot"
#| fig-alt: "A line plot on a polar axis"
ggmice(dat, aes(popular, teachpop)) + 
  geom_jitter()
```

@@

```{R, eval=FALSE}
#| layout-ncol: 2
#| echo: false
# plot_pattern(dat)
# plot_corr(dat)
```

@@

```{R}
#| prompt: true
meth <- make.method(dat)
meth
```

```{R}
#| prompt: true
pred <- quickpred(dat)
pred
```

Adjust the methods vector.
```{R}
#| prompt: true
meth["popular"] <- "2l.pmm"
```

Adjust the predictor matrix.
```{R}
#| prompt: true
pred["popular", "school"] <- -2
pred["popular", "sex"] <- 2
```

Visualize the imputation methods and predictors.
```{R}
plot_pred(pred, method = meth)
```

Impute the data.
```{R}
#| prompt: true
imp <- mice(
  data = dat,
  method = meth,
  predictorMatrix = pred,
  printFlag = FALSE
)
```

Evaluate the convergence.
```{R}
#| prompt: true
plot_trace(imp)
```

Evaluate the distribution of imputed values.
```{R}
ggmice(imp, aes(popular, group = .imp)) + 
  geom_density() 
```

Evaluate the distribution of imputed values.
```{R}
ggmice(imp, aes(.imp, popular)) + 
  geom_jitter(alpha = 0.05) +
    geom_boxplot()
```

```{R}
ggmice(imp, aes(popular, teachpop)) + 
  geom_jitter() +
  facet_wrap(~ .imp)
```

Analyze the imputed data.
```{R}
fit <- with(
  imp,
  lmer(teachpop ~ popular + texp + (1 | school))
)
```

Pool the estimates.
```{R}
pool(fit)
```

Display results in table.
```{R}
testEstimates(as.mitml.result(fit), extra.pars = TRUE)
```

### 2023-12-07 docs/manuscript.tex
Add scatterplots

\begin{tcolorbox}[enhanced jigsaw, arc=.35mm, breakable, leftrule=.75mm, colback=white, bottomrule=.15mm, toprule=.15mm, left=2mm, rightrule=.15mm, opacityback=0]

@@

\begin{verbatim}
R> library(mice)
R> library(ggmice)
\end{verbatim}

@@

\begin{verbatim}
R> dat <- popmis[, c("school", "popular", "sex")] 
\end{verbatim}

@@

\begin{minipage}[t]{0.50\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{manuscript_files/figure-pdf/unnamed-chunk-5-1.pdf}

\end{minipage}%
%
\begin{minipage}[t]{0.50\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{manuscript_files/figure-pdf/unnamed-chunk-5-2.pdf}

\end{minipage}%

@@

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-6-1.pdf}

@@

\begin{verbatim}
R> meth <- make.method(dat)
\end{verbatim}

\begin{verbatim}
R> pred <- quickpred(dat)
R> plot_pred(pred, method = meth)
\end{verbatim}

@@

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-8-1.pdf}

@@

@@

@@

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-10-1.pdf}

@@

>>>>

\begin{tcolorbox}[enhanced jigsaw, left=2mm, leftrule=.75mm, bottomrule=.15mm, toprule=.15mm, breakable, colback=white, opacityback=0, arc=.35mm, rightrule=.15mm]

@@

\begin{verbatim}
R> set.seed(123)
R> library(mice)
R> library(ggmice)
R> library(ggplot2)
R> library(miceadds)
R> library(lme4)
R> library(mitml)
R> library(broom.mixed)
\end{verbatim}

@@

\begin{verbatim}
R> dat <- popmis[, c("school", "teachpop", "popular", "texp", "sex")] 
\end{verbatim}

\begin{verbatim}
ggmice(dat, aes(popular, teachpop)) + 
  geom_jitter()
\end{verbatim}

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-4-1.pdf}

\caption{Polar axis plot}

@@

@@

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-7-1.pdf}

@@

\begin{verbatim}
R> meth <- make.method(dat)
R> meth
\end{verbatim}

\begin{verbatim}
  school teachpop  popular     texp      sex 
      ""       ""    "pmm"       ""       "" 
\end{verbatim}

\begin{verbatim}
R> pred <- quickpred(dat)
R> pred
\end{verbatim}

\begin{verbatim}
         school teachpop popular texp sex
school        0        0       0    0   0
teachpop      0        0       0    0   0
popular       0        1       0    1   1
texp          0        0       0    0   0
sex           0        0       0    0   0
\end{verbatim}

Adjust the methods vector.

\begin{verbatim}
R> meth["popular"] <- "2l.pmm"
\end{verbatim}

Adjust the predictor matrix.

\begin{verbatim}
R> pred["popular", "school"] <- -2
R> pred["popular", "sex"] <- 2
\end{verbatim}

Visualize the imputation methods and predictors.

\begin{verbatim}
plot_pred(pred, method = meth)
\end{verbatim}

@@

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-12-1.pdf}

@@

Impute the data.

@@

Evaluate the convergence.

@@

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-14-1.pdf}

@@

Evaluate the distribution of imputed values.

\begin{verbatim}
ggmice(imp, aes(popular, group = .imp)) + 
  geom_density() 
\end{verbatim}

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-15-1.pdf}

Evaluate the distribution of imputed values.

\begin{verbatim}
ggmice(imp, aes(.imp, popular)) + 
  geom_jitter(alpha = 0.05) +
    geom_boxplot()
\end{verbatim}

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-16-1.pdf}

\begin{verbatim}
ggmice(imp, aes(popular, teachpop)) + 
  geom_jitter() +
  facet_wrap(~ .imp)
\end{verbatim}

{\centering \includegraphics{manuscript_files/figure-pdf/unnamed-chunk-17-1.pdf}

Analyze the imputed data.

\begin{verbatim}
fit <- with(
  imp,
  lmer(teachpop ~ popular + texp + (1 | school))
)
\end{verbatim}

Pool the estimates.

\begin{verbatim}
pool(fit)
\end{verbatim}

\begin{verbatim}
Class: mipo    m = 5 
         term m  estimate         ubar            b            t dfcom
1 (Intercept) 5 2.4091354 2.241304e-02 1.712964e-03 2.446860e-02  1995
2     popular 5 0.2597284 2.353344e-04 1.209648e-04 3.804922e-04  1995
3        texp 5 0.0484727 7.728295e-05 3.236252e-06 8.116646e-05  1995
         df        riv     lambda        fmi
1 432.50579 0.09171257 0.08400798 0.08821454
2  26.88403 0.61681474 0.38149995 0.42289329
3 909.68447 0.05025044 0.04784615 0.04993264
\end{verbatim}

Display results in table.

\begin{verbatim}
testEstimates(as.mitml.result(fit), extra.pars = TRUE)
\end{verbatim}

\begin{verbatim}

Call:

testEstimates(model = as.mitml.result(fit), extra.pars = TRUE)

Final parameter estimates and inferences obtained from 5 imputed data sets.

             Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
(Intercept)     2.409     0.156    15.401   566.786     0.000     0.092     0.087 
popular         0.260     0.020    13.315    27.483     0.000     0.617     0.422 
texp            0.048     0.009     5.380  1747.294     0.000     0.050     0.049

                            Estimate 
Intercept~~Intercept|school    0.310 
Residual~~Residual             0.307 
ICC|school                     0.502

Unadjusted hypothesis test as appropriate in larger samples.
\end{verbatim}
